Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-17 model.tex
finished rewriting all but the MASE section of model.tex

A random-walk predictor simply uses the last observed measurement as
the forecast: that is, the value $px_i$ at time $i$ is predicted using
the relation $$px_i = x_{i-1}$$ The \naive ~prediction strategy
averages the prior observations to generate the forecast: $$px_i =
\sum_{j=1}^{i-1}\frac{x_j}{i-1}$$ While both of these methods are
simplistic, they are not without merit.  For a time series near the
high end of the complexity spectrum---i.e., one that possesses very
little predictive structure---these two methods can actually be the
best choice.  In forecasting currency exchange rates, for instance,
sophisticated econometrics-based prediction models fail to
consistently outperform the random-walk method~\cite{rwMeese,rwCCE}.
These signals are volatile, noisy, and possess very little predictive
structure, but they do not---on the average---vary a great deal, so
the random-walk method's strategy of simply guessing the last known
value is not a bad choice.  If a signal has a unimodal distribution
with low variance, the \naive ~prediction strategy will perform quite
well---even if the signal is highly complex---simply because the mean
is a good approximation of the future behavior.  Moreover, a temporal
average effects a low-pass filtering operation.  In a signal with very
little predictive structure, then, the \naive ~prediction strategy
effectively mitigates the complexity.

@@

Perhaps the simplest way to capture and exploit the structure of data
is to fit a hyperplane to the dataset and then use that it to make
predictions.  The roots of this approach date back to Yule's 1927
invention of the autoregressive schema~\cite{weigend93}, which
forecasts the next time step through a weighted average of past
observations: $$px_i = \sum a_i x_i$$ The weighting coefficients $a_i$
are computed using a polynomial of degree $p$ that is fit to the past
$p+1$ values of $x_i$.  To account for noise in the data, one can add
a so-called ``moving average'' term to the model; to remove
nonstationarities, one can detrend the data using a differencing
operation.  A strategy that incorporates all three of these features
is called a \emph{nonseasonal ARIMA model}.  If periodic structure is
present in the data, a \emph{seasonal ARIMA model}, which adds a
sampling operation that filters out periodicities, can be a good
choice.

@@

ARIMA forecasting is a common and time-tested procedure.  Its
adjustments for seasonality, nonstationarity, and noise make it an
appropriate choice for short-term predictions of time-series data
generated by a wide range of processes.  If information is being
generated and/or transmitted in a nonlinear way, however, a global
linear fit is inappropriate and ARIMA forecasts will be inaccurate.
Another weakness of this method is prediction horizon, since an ARIMA
forecast is guaranteed to converge to the mean after some number of
predictions, depending on model order.  To avoid this issue, we build
forecasts in a stepwise fashion: i.e., fit the model to the existing
data, use it to perform a one-step prediction, rebuild the model using
the latest observations, and iterate until the desired prediction
horizon is reached.  (For consistency, we take the same approach with
the other three models in this study as well, even though that amounts
to artificially hobbling LMA.)

\subsection{Nonlinear Forecasting: The Lorenz Method of Analogues}
\label{sec:lma}

When a generating process exhibits deterministic predictive nonlinear
structure a common approach is to reconstruct the generating processes
dynamics using a technique called delay-coordinate embedding and then
use the geometric structure of this reconstruction as a proxy for the
information transmission occurring in the underlying process to
predict the future. In this section we discuss the theory and
implementation of this forecasting scheme.

 \subsubsection{Reconstructing hidden dynamics {\color{blue} EDITABLE}}

Delay-coordinate embedding allows one to reconstruct a dynamical system's full
state-space dynamics from a \emph{single} scalar time-series
measurement---provided that some conditions hold regarding that data.
Specifically, if the underlying dynamics and the measurement
function---the mapping from the unknown state vector $\vec{X}$ to the
scalar value $X_{i,obs}$ that one is measuring---are both smooth and generic,
Takens~\cite{takens} formally proves that the delay-coordinate map
\[
F(\tau,m)(\vec{X}) = ([X_{i+\tau,obs} ~ X_{i+\tau,obs} ~ \dots ~X_{i+m\tau,obs}])
\]
from a $d$-dimensional smooth compact manifold $M$ to $\mathbb{R}^{2d+1}$,
where $t$ is time, is a diffeomorphism on $M$---in other words, that
the reconstructed dynamics and the true (hidden) dynamics have the
same topology.

This is an extremely powerful result: among other things, it means
that one can build a formal model of the full system dynamics without
measuring (or even knowing) every one of its state variables.  This is
the foundation of this modeling approach.
The first step in the process is to estimate values for the two free
parameters in the delay-coordinate map: the delay $\tau$ and the
dimension $m$.  We follow standard procedures for this, choosing the
first minimum in the average mutual information as an estimate of
$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
method of \cite{KBA92} to estimate $m$.  A
plot of the data from Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored}, embedded following this
procedure, are shown in Figure~\ref{fig:embedding}.

 \begin{figure}
   \centering
\begin{subfigure}{\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colipc3d.png}
    \caption{\col }
    \label{fig:colEmbedding}
  \end{subfigure}%  \\

@@

     %\includegraphics[width=\textwidth]{figs/colipc3d.png}
     \caption{A 3D projection of the delay-coordinate embeddings of the traces
 from: (a) \col seen in Figure~\ref{fig:col-ts} (b)\gcc seen in Figure~\ref{fig:gcc-ts} and (c) \svdfive seen as the fifth regime in Figure~\ref{fig:svd-ts-colored}.
 }
 \label{fig:embedding}
 \end{figure}

%% can cut for space if need be:
The coordinates of each point in these plots are differently delayed
elements of the instructions per cycle time series
$X_{i,obs}$: that is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
$X_{i+2\tau,obs}$ on the third, and so on.
Structure in these kinds of plots---clearly visible in
Figure~\ref{fig:embedding} (a) is an indication of
structure in the information transmission process, that can be leveraged for prediction. Where the water becomes muddy is with embedding like those in Figure~\ref{fig:embedding} (b) \& (c). Figure~\ref{fig:embedding} (b) shows little to no structure, this embedding is essentially a noise cloud; with an embedding like this we would expect the forecast accuracy that uses this structure to be lacking. However, with Figure~\ref{fig:embedding} (c) more structure is present but the structure is surrounded in noise. Is the structure enough to aid in accurately forecasting the signal? This is a difficult question to answer and one that comes back to the main contributions of this paper. It may be that \gcc has some redundancy but that structure is not amenable to this method, similarly \svdfive clearly has structure but it is obfuscated by noise. By quantifying the balance between normalized redundancy (predictive structure) and entropy for these real valued time series, we can begin to answer these questions.  If enough (redundancy) predictive structure is present in these embeddings as seen in Figure~\ref{fig:embedding} (a) and possibly in (c), that structure can be used to build a forecast model which we discuss in the next section.
 \subsubsection{LMA: Using dynamics in forecasting {\color{blue} EDITABLE}}

Given a nonlinear model of a deterministic dynamical system in the
form of a delay-coordinate embedding like those in Figure~\ref{fig:embedding},
one can build deterministic forecast algorithms by capturing and
exploiting the geometry of the embedding.  Many techniques have been
developed by the dynamical systems community for this purpose
(e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).  Perhaps the most straightforward
is the ``Lorenz method of analogues'' (LMA), which is essentially
nearest-neighbor prediction in the embedded state
space~\cite{lorenz-analogues}.  Even this simple algorithm---which
builds predictions by finding the nearest neighbor in the embedded space of the given point, then taking that neighbor's path as the
forecast, provides accurate forecasts when the generating process is a deterministic dynamical system. In particular, when an observation in a time series has redundancy in the form of nonlinear predictive structure this forecast scheme should be a great choice. However, since this method is based on near-neighbor prediction in a reconstructed state-space, if a system does not exhibit deterministic structure this method can do quite poorly. {\color{red}[[Joshua:Not a fan of this wording]]}
%% [[If space: Add a segue sentence: next section uses permutation
%% entropy to explore that conjecture.]]

%\subsection{Comparing Prediction Accuracy}

\subsection{Prediction Accuracy: Mean Absolute Scaled Error (MASE) {\color{blue} EDITABLE}}
\label{sec:accuracy}

In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$. The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:model}. The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary. In particular, we perform $k$ one-step predictions, after each one-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction. This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.

As a figure of merit we calculate the Mean Absolute Scaled Error (MASE)\cite{MASE} between the true and predicted signals: 
%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
The scaling term for MASE:
$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$. This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." We chose MASE because it allows for fair comparison across varying methods, prediction horizons and signal scales. When a forecast results in a $MASE<1$ the prediction method produced, on average, smaller errors than the one-step errors from the in-sample random walk forecast strategy. Analogously, $MASE>1$ means that the prediction method did worse, on average than the one-step errors for the in-sample random walk forecast strategy.

>>>>

A random-walk predictor simply uses the last observed measurement as
the forecast: that is, the value $px_i$ at time $i$ is predicted using
the relation $$px_i = x_{i-1}$$ The prediction strategy that we refer
to using the term ``\naive'' averages the prior observations to
generate the forecast: $$px_i = \sum_{j=1}^{i-1}\frac{x_j}{i-1}$$
While both of these methods are simplistic, they are not without
merit.  For a time series near the high end of the complexity
spectrum---i.e., one that possesses very little predictive
structure---these two methods can actually be the best choice.  In
forecasting currency exchange rates, for instance, sophisticated
econometrics-based prediction models fail to consistently outperform
the random-walk method~\cite{rwMeese,rwCCE}.  These signals are
volatile, noisy, and possess very little predictive structure, but
they do not---on the average---vary a great deal, so the random-walk
method's strategy of simply guessing the last known value is not a bad
choice.  If a signal has a unimodal distribution with low variance,
the \naive ~prediction strategy will perform quite well---even if the
signal is highly complex---simply because the mean is a good
approximation of the future behavior.  Moreover, a temporal average
effects a low-pass filtering operation.  In a signal with very little
predictive structure, then, the \naive ~prediction strategy
effectively mitigates the complexity.

@@

A simple and yet powerful way to capture and exploit the structure of
data is to fit a hyperplane to the dataset and then use that it to
make predictions.  The roots of this approach date back to Yule's 1927
invention of the autoregressive schema~\cite{weigend93}, which
forecasts the next time step through a weighted average of past
observations: $$px_i = \sum a_i x_i$$ The weighting coefficients $a_i$
are computed using a polynomial of degree $p$ that is fit to the past
$p+1$ values of $x_i$.  To account for noise in the data, one can add
a so-called ``moving average'' term to the model; to remove
nonstationarities, one can detrend the data using a differencing
operation.  A strategy that incorporates all three of these features
is called a \emph{nonseasonal ARIMA model}.  If evidence of periodic
structure is present in the data, a \emph{seasonal ARIMA model}, which
adds a sampling operation that filters out periodicities, can be a
good choice.

@@

ARIMA forecasting is a common and time-tested procedure.  Its
adjustments for seasonality, nonstationarity, and noise make it an
appropriate choice for short-term predictions of time-series data
generated by a wide range of processes.  If information is being
generated and/or transmitted in a nonlinear way, however, a global
linear fit is inappropriate and ARIMA forecasts will be inaccurate.
Another weakness of this method is prediction horizon, since an ARIMA
forecast is guaranteed to converge to the mean after some number of
predictions, depending on model order.  To avoid this issue, we build
forecasts in a stepwise fashion: i.e., fit the model to the existing
data, use it to perform a one-step prediction, rebuild the model using
the latest observations, and iterate until the desired prediction
horizon is reached.  (For consistency, we take the same approach with
the other three models in this study as well, even though this amounts
to artificially hobbling LMA.)

\subsection{A Nonlinear Prediction Strategy}
\label{sec:lma}

When the temporal progressions in a time series are produced by a
deterministic nonlinear process, one can use a technique called
delay-coordinate embedding
%
%% Note: claiming that we can reconstruct the dynamics of the
%% underlying generating process isn't right.  That would be
%% equivalent to solving the system identification problem!  We're
%% not reconstructing the underlying system, just its output.
%
to model the structure of the information generation and transmission
occurring in the underlying process, then use that reconstruction to
generate forecasts.  This section discusses the theory and
implementation of a prediction strategy that is based on this idea.

Delay-coordinate embedding allows one to reconstruct a dynamical
system's full state-space dynamics from a scalar time-series
measurement---provided that some conditions hold regarding that data.
Specifically, if the underlying dynamics and the measurement
function---the mapping from the unknown state vector $\vec{X}$ to the
observed value $x_i$---are both smooth and generic,
Takens~\cite{takens} formally proves that the delay-coordinate map
\[
F(\tau,m)(\vec{X}) = ([x_{i} ~ x_{i+\tau} ~ \dots ~x_{i+m\tau}])
\]
from a $d$-dimensional smooth compact manifold $M$ to
$\mathbb{R}^{2d+1}$ is a diffeomorphism on $M$---in other words, that
the reconstructed dynamics and the true (hidden) dynamics have the
same topology.  This is an extremely powerful result: among other
things, it means that one can model the full system dynamics without
measuring, or even knowing, every one of its state variables.

The first step in the delay-coordinate embedding process is to
estimate values for the two free parameters in the map: the delay
$\tau$ and the dimension $m$.  We follow standard procedures for this,
choosing the first minimum in the average mutual information as an
estimate of $\tau$~\cite{fraser-swinney} and using the
false-near-neighbor method of~\cite{KBA92} to estimate $m$.  Plots of
the data from Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored},
embedded following this procedure, are shown in
Figure~\ref{fig:embedding}.
 \begin{figure}
   \centering
\begin{subfigure}{\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colipc3d.png}
    \caption{\col }
    \label{fig:colEmbedding}
  \end{subfigure}%  \\

@@

     %\includegraphics[width=\textwidth]{figs/colipc3d.png}
     \caption{3D projections of delay-coordinate embeddings of the
       traces from (a) Figure~\ref{fig:col-ts} (b)
       Figure~\ref{fig:gcc-ts} and (c) the fifth (green) segment of
       Figure~\ref{fig:svd-ts-colored}.}
 \label{fig:embedding}
 \end{figure}
%% can cut for space if need be:
%The coordinates of each point in these plots are differently delayed
%elements of the instructions per cycle time series $X_{i,obs}$: that
%is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
%$X_{i+2\tau,obs}$ on the third, and so on.  
%

Geometric structure in these kinds of plots is an indication of
structure in the underlying information generation/transmission
process.  The dynamical systems community has developed a number of
methods that leverage this structure to generate predictions
(e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).  One of the
most straightforward of these is the ``Lorenz method of analogues''
(LMA), which is essentially nearest-neighbor prediction in the
embedded\footnote{Lorenz's original formulation used the full system
  state space; the method was later extended to embedded dynamics by
  Kantz~\cite{kantz-lma-citation}.} space~\cite{lorenz-analogues}.

Even this simple algorithm---which builds predictions by finding the
nearest neighbor in the embedded space of the given point, then taking
that neighbor's path as the forecast---provides accurate forecasts
when the generating process is a deterministic dynamical system.
Since LMA does not rest on an assumption of linearity, as ARIMA does,
it can be applied to data produced by both linear and nonlinear
processes.  If the underlying generating process is not deterministic,
however, it can perform poorly.  In Figure~\ref{fig:embedding}(b), for
instance, little structure is visible, so LMA may not work well.  More
structure appears to be present in~Figure~\ref{fig:embedding}(c), but
this reconstruction also appears to contain some noise.

The question as to how much structure is present in a
reconstruction---and how much of that structure can be captured and
used by LMA---is apropos of the central question treated in this
paper.  It may be that \gcc has some redundancy that LMA cannot
exploit, or that the structure in \svdfive is effectively obfuscated,
from the standpoint of the LMA method, by noise.  By quantifying the
balance between normalized redundancy (that is, predictive structure)
and entropy for these real-valued time series, we can begin to answer
these questions.

% If enough predictive structure is present in these embeddings, that
% structure can be used to build an effective forecast model.

\subsection{Assessing Prediction Accuracy}
\label{sec:accuracy}

To study the relationship between predictability and complexity, we
use these four methods to generate predictions of all 360 traces
described in Section~\ref{sec:methods}, then calculate the error
between the predictions and the true continuations.  Specifically, we
split each time series into two pieces: the first 90\%, referred to as
the ``initial training" signal and denoted $\{x_i\}_{i=1}^{n}$, and
the last 10\% known as the ``test" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
The initial training signal is used to build the model, following the
procedures described in the previous section; that model is used to
generate a prediction of the value of $x_{n+1}$, which is then
compared to the true continuation, $c_{n+1}$.  The model is then
rebuilt using $\{x_i\}_{i=1}^{n+1}$ and the process repeats $k$ times,
out to the end of the observed time series.  This ``one step
prediction'' process is not technically necessary in the LMA method,
which can generate arbitrary-length predictions, but the performance
of the other three methods used here will degrade if the associated
models are not periodically rebuilt.  In order to make the comparison
fair, we used an iterative one-step prediction schema {\sl for all
  four methods}.  This has the slightly confusing effect of causing
the ``test'' signal to be used both to assess the accuracy of each
model and for periodic refitting.

To assess prediction accurary, we calculate the mean absolute scaled
error (MASE) between the true and predicted signals:
% 
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
% 
This error method was introduced in \cite{MASE} as a ``generally
applicable measurement of forecast accuracy without the problems seen
in the other measurements."  Note that the scaling term in the
denominator
% 
% of the MASE value
% $$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
% 
is the average in-sample forecast error for a random-walk prediction
$px_i=x_{i-1}$ over a signal of length $k$.  That is, MASE is a
normalized measure.  A MASE value below one means that the prediction
error was, on the average, smaller than the error of a one-step
in-sample random-walk forecast on the same data.  Analogously, a MASE
value greater than one means that the corresponding prediction method
did \emph{worse}, on average, than a one-step in-sample random-walk
method.  We chose this error metric because it allows for fair
comparison across varying methods, prediction horizons, and signal
scales.
